\documentclass[sigconf, nonacm]{acmart}

\usepackage{booktabs}

%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[Explainable and Trustworthy AI - Project]{}{}{}

\begin{document}

\title{Unsupervised Concept Discovery for Medical Vision--Language Models:\\A Rigorous Characterization of Sparse-Autoencoder Failure and Deterministic Alternatives}

\author{Marc'Antonio Lopez}
\affiliation{%
  \institution{Politecnico di Torino}
  \city{Turin}
  \country{Italy}
}

\author{Nicolò Colle}
\affiliation{%
  \institution{Politecnico di Torino}
  \city{Turin}
  \country{Italy}
}

\author{Name Surname}
\affiliation{%
  \institution{Politecnico di Torino}
  \city{Turin}
  \country{Italy}
}

\begin{abstract}
Medical Vision--Language Models (VLMs) are accurate but opaque, and concept-based explainability promises to expose their internal knowledge as human-readable units. We re-implement the MedConcept pipeline~\cite{haque2026medconcept}---sparse autoencoder (SAE) decomposition of a frozen VLM's image embeddings, alignment to a curated radiology vocabulary, and LLM-as-judge evaluation---on BiomedCLIP~\cite{zhang2023biomedclip}, on \emph{two} scales: IU X-Ray~\cite{demner2016iuxray} (${\sim}7$k images) and ROCOv2 (${\sim}80$k, a $10{\times}$ scale test). Our first result is methodological: the slot-wise cross-seed Jaccard used as the stability headline is \emph{floor-by-construction}---a single SAE whose features are relabelled by a random bijection scores ${\approx}0.0077$, indistinguishable from real cross-seed---so it cannot establish non-identifiability. We therefore report a permutation-invariant \emph{matched} metric against a subspace-conditioned (effective-rank) null. Under this honest metric, SAE concepts are genuinely non-identifiable: best-match cosine $0.30$--$0.33$ at only $1.67$--$1.81{\times}$ the conditioned null, with ${\approx}0\%$ of features matching at $\geq0.9$ across all seed pairs, and naming cosine capped at $0.40$--$0.63$ ($0\%$ above $0.7$) even with a chest-specific vocabulary. Crucially, the $10{\times}$ ROCOv2 scale does \emph{not} lift identifiability (matched ratio $1.67{\times}{\to}1.81{\times}$, frac${\geq}0.9$ $0\%{\to}0.3\%$), refuting the data-scale hypothesis and localising the binding constraint to the \emph{representation site}: SAE-on-CLS decomposes an already-compressed, contrastive-shaped global vector, and no dictionary over $30$--$500{\times}$ the input's intrinsic dimension is identifiable. SPLiCE~\cite{bhalla2024spliece}, a deterministic decomposition over a fixed dictionary, sidesteps the estimation problem and is the sole judge-ready method. We report these outcomes honestly---including the slot-wise metric artefact and a pending (machinery-complete, inputs-regenerated) LLM-judge run---as a reproducibility study of \emph{where} unsupervised concept discovery breaks and \emph{why}.
\end{abstract}

\keywords{Concept-based explainability, sparse autoencoders, medical vision--language models, interpretability evaluation, reproducibility}

\maketitle

\section{Introduction}
Medical VLMs attain strong performance on pathology classification, segmentation, and report generation, yet their internal representations are high-dimensional, polysemic, and opaque---a serious concern in safety-critical clinical use. Classical post-hoc tools such as saliency maps and attention visualisations produce local explanations that seldom reveal the semantic structure the model has actually learned.

Concept-based explainability reframes an explanation in terms of human-interpretable units. Where supervised approaches such as Concept Bottleneck Models~\cite{koh2020concept} and TCAV~\cite{kim2018tcav} require a pre-defined, labelled concept set, recent work attempts \emph{unsupervised} discovery: decomposing a frozen model's representations into sparse features, naming each feature against a clinical vocabulary, and validating the result quantitatively. MedConcept~\cite{haque2026medconcept} instantiates this as a three-stage pipeline (sparse-activation extraction, vocabulary alignment, LLM-judge evaluation) and introduces the Aligned/Unaligned/Uncertain metrics for quantitative interpretability assessment.

A central but under-reported question is whether the concepts such pipelines discover are \emph{reproducible} and \emph{clinically valid}. In this work we re-implement the MedConcept paradigm faithfully, and find---rigorously, and against analytical nulls---that the learned sparse decomposition is essentially non-identifiable: independently seeded SAEs share no canonical features (a permutation-invariant matched metric lands at $1.67$--$1.81{\times}$ a subspace-conditioned null, with ${\approx}0\%$ of features matching at $\geq0.9$ cosine), the naming signal is capped at $0.40$--$0.63$ cosine, and a large fraction of features never fire. Crucially, we first expose that the slot-wise index Jaccard used as the stability headline is \emph{floor-by-construction} (a relabelled copy of one SAE scores ${\approx}0.0077$ on it), so non-identifiability must be measured permutation-invariantly.

We then investigate three directions intended to address the root causes, and an organisational extension:
\begin{itemize}
  \item \textbf{Baseline} (512-d projected SAE): a faithful re-implementation of MedConcept, characterised as a documented failure case.
  \item \textbf{Path A} (768-d pre-projection hidden state SAE): targets the representation-location mismatch hypothesised to drive non-identifiability.
  \item \textbf{Path B} (SPLiCE): replaces the learned dictionary with a fixed RadLex dictionary, removing the estimation problem entirely.
  \item \textbf{Extension} (concept organisation): clusters flat concept lists into structured, hierarchy-grounded families.
\end{itemize}

\noindent\textbf{Contributions.} (1)~A \emph{relabeling control} proving the slot-wise cross-seed Jaccard is floor-by-construction: one SAE with its features renamed by a random bijection scores ${\approx}0.0077$ on it, yet $1.0$ on a permutation-invariant matched metric---so the slot-wise number measures feature \emph{indexing}, not feature \emph{identity}, and cannot establish non-identifiability. (2)~A permutation-invariant \emph{matched-stability} metric with a \emph{subspace-conditioned} (effective-rank) null that controls for data-manifold concentration; under it, SAE concepts are genuinely non-identifiable (best-match $1.67$--$1.81{\times}$ the null, ${\approx}0\%$ at $\geq0.9$). (3)~A direct test of the data-scale hypothesis: ROCOv2 at $10{\times}$ IU X-Ray scale does \emph{not} restore identifiability, localising the binding constraint to the representation site (SAE-on-CLS), not data volume. (4)~An honest, partially-negative evaluation---including the slot-wise artefact and a machinery-complete, inputs-regenerated judge run pending a converged GPU execution---rather than a curated success story.

\section{Related work}
\textbf{Concept-based explainability.} Concept Bottleneck Models~\cite{koh2020concept} restructure a network to predict an explicit concept layer before the classifier, trading minor accuracy for auditable, concept-level intervention. TCAV~\cite{kim2018tcav} tests directional sensitivity of a logit to user-defined Concept Activation Vectors. Both presuppose supervised, pre-defined concepts---precisely the limitation unsupervised discovery targets.

\textbf{Sparse autoencoders for monosemantic decomposition.} Bricken et al.~\cite{bricken2023monosemanticity} showed that overcomplete dictionary-learning autoencoders trained on transformer activations yield individual latents that activate on coherent, human-nameable features. Gao et al.~\cite{gao2024scaling} introduced the k-sparse (Top-K) autoencoder, which fixes sparsity exactly via top-$k$ gating, eliminates the reconstruction--sparsity trade-off, and reduces dead latents; we adopt this architecture.

\textbf{SAE stability and universality.} A growing line asks whether SAE features are canonical. Lan et al.~\cite{lan2024universality} showed that SAE feature spaces across language models are similar only up to rotation, sharing a subspace rather than identical features. Leask et al.~\cite{leask2025canonical} argued SAEs are simultaneously incomplete and non-atomic, and therefore ``do not find canonical units of analysis.'' This weak-vs-strong universality distinction motivates our matched-stability metric: we no longer expect identical features across seeds, only a shared subspace.

\textbf{Sparse concept decomposition.} SPLiCE~\cite{bhalla2024spliece} bypasses a learned dictionary and decomposes a frozen CLIP embedding directly into a sparse, non-negative combination over a fixed, interpretable concept dictionary. Because the dictionary is fixed and the method deterministic, it sidesteps the non-identifiability of learned factorisation---a property we exploit as a data-frugal baseline.

\textbf{Medical VLMs and the reference framework.} BiomedCLIP~\cite{zhang2023biomedclip} is a domain-specific contrastive VLM (ViT image encoder + PubMedBERT text encoder) pretrained on PMC-15M; it supplies both the shared 512-d space and the pre-projection 768-d hidden states our pipeline decomposes. MedConcept~\cite{haque2026medconcept} provides the methodological template---sparse extraction, vocabulary alignment, and LLM-judge evaluation with the Aligned/Unaligned/Uncertain metrics---which we re-implement and stress-test. RadLex~\cite{langlotz2006radlex} provides the curated radiology lexicon used for naming and organisation.

\section{Research gaps}
From the literature and the project brief we identify six gaps that this work addresses:
\begin{enumerate}
  \item \textbf{Concept instability / non-identifiability.} Learned sparse factorisations are not guaranteed to be reproducible across seeds, and in the small-data regime the decomposition is provably non-identifiable. No prior medical-VLM concept work we are aware of reports seed-stability against an analytical null.
  \item \textbf{Clinical validity of discovered concepts.} A cosine-assigned name is cosmetic unless it tracks real pathology; concepts are rarely validated against held-out ground-truth labels.
  \item \textbf{Representation-location mismatch.} The SAE-interpretability literature operates on raw hidden states, yet reference pipelines fit the SAE on the already-projected, modality-gap-bearing CLIP space---a structurally poorer target.
  \item \textbf{Small-data robustness vs.\ learned dictionaries.} Learned SAEs require $10^5$--$10^6$ activations; medical corpora are 2--3 orders of magnitude smaller, yielding large dead-feature fractions.
  \item \textbf{Concepts treated as independent.} Concept explanations are typically flat top-$k$ lists, ignoring natural anatomical/hierarchical structure.
\end{enumerate}

\section{Methodology}
\subsection{Backbone, vocabulary, and split}
We use BiomedCLIP (ViT-B/16 image encoder with a 768-d hidden state, projected by a frozen $\mathrm{visual\_projection}: \mathbb{R}^{768}\!\to\!\mathbb{R}^{512}$ into a shared contrastive space). Images are resized to $224{\times}224$; both image and text embeddings are L2-normalised at extraction. The concept vocabulary is built from RadLex: terms are encoded with BiomedCLIP's text encoder, ranked by cosine similarity to a multi-centroid relevance score computed from 39 anchor queries across 13 clinical sub-domains, and the top terms plus 14 NIH ChestX-ray14 seed terms are retained, yielding ${\sim}1{,}030$ chest-radiography-relevant terms cached as 512-d text embeddings.

The shared 512-d space exhibits a \emph{modality gap}~\cite{liang2022mindthegap}: a systematic offset between the image and text cones (cross-modal image--text cosine ${\approx}0.27$, centroid L2 distance ${\approx}0.95$). Following prior work we estimate it as the difference of the mean image and vocabulary embeddings and subtract it before cosine naming/decomposition.

All experiments use \emph{two} chest-radiography corpora. IU X-Ray~\cite{demner2016iuxray} (7,470 frontal/lateral images, 3,955 reports) is the primary, small-scale setting; ROCOv2 (${\sim}80$k images) is a $10{\times}$-scale replication that lets us test the data-volume hypothesis directly.  In order to avoid data leakage, we split IU X-Ray \emph{by radiographic study} (IU key $=\texttt{patient\_IM-study}$), so no IU study appears at the same time in both train and test sets; in opposite, ROCOv2 is simply split by image, since there is no shared patient structure to leak. The IU split yields 5,955 train / 1,515 test images (3,081 / 771 studies) with verified group overlap $=0$, recomputed deterministically on every run (sorted study set, fixed seed).

\subsection{Baseline: Top-K SAE on the 512-d projected space}
We train a Top-K SAE~\cite{gao2024scaling} ($\mathrm{ReLU}(W_{\mathrm{enc}}(x-b_{\mathrm{dec}})+b_{\mathrm{enc}})$, hard top-$k$ sparsification, $b_{\mathrm{dec}}$ initialised to the geometric median of training data) on the 512-d image embeddings. Configuration: dictionary size $D{=}2048$, $k{=}32$, 8,000 steps, $\mathrm{lr}{=}5{\times}10^{-5}$, batch size 256, five seeds $(0,42,123,456,789)$. Concept naming assigns each \emph{live} (non-dead) decoder row to its closest RadLex term by cosine similarity, after gap correction ($W_{\mathrm{dec}}\!\leftarrow\!W_{\mathrm{dec}}-\mathrm{gap}$). Per-image pseudo-reports list the top-5 active concepts with a stated dominant concept.

\subsection{Path A: Top-K SAE on the 768-d hidden state}
Path A addresses Gap~3. Instead of $\mathrm{visual\_projection}(\cdot)$, we extract the 768-d CLS token \emph{before} the projection and train an identical Top-K SAE on these raw (un-normalised) hidden states. Because RadLex text lives in 512-d, we bridge the two spaces with the \emph{frozen} projection: $\hat W_{\mathrm{dec}}^{512}=W_{\mathrm{dec}}^{768}W_{\mathrm{proj}}^{\top}$, gap-correct $\hat W_{\mathrm{dec}}^{512}$, then name against RadLex as in the baseline. Dead features are detected pre-projection (a zero 768-d row projects to $-\mathrm{gap}$ with non-zero norm) and cross-checked with an activation-based mask.

\subsection{Path B: SPLiCE on the RadLex dictionary}
Path B addresses Gaps~1 and~4. We decompose each gap-corrected 512-d image embedding into a sparse, non-negative combination over the fixed RadLex text dictionary using Orthogonal Matching Pursuit to select $k{=}32$ atoms, followed by a non-negative least-squares re-solve on the selected support. The dictionary is never estimated from the ${\sim}6$k samples, so the method is deterministic by construction and has no cross-seed instability. The output schema mirrors the SAE explanations so the same judge can score it.

\subsection{Concept-organisation extension}
Addressing Gap~5, we cluster the active vocabulary into concept families by agglomerative clustering (average linkage, cosine metric). Each cluster is annotated with the most specific shared RadLex anatomical ancestor, with two degeneracy guards: leaf-root rejection and a \emph{subtree-size canopy} that rejects candidate ancestors whose descendant count exceeds 1\% of the ontology (e.g.\ the trivially generic ``anatomical entity'' node), so labels are honest rather than vacuously generic. The stage emits structured per-image families plus organisation metrics (silhouette, redundancy reduction, RadLex coverage). It is standalone and does not feed the judge.

\subsection{Evaluation metrics}
\textbf{Stability (matched, the headline).} For each decoder row in seed~$i$ we take the best cosine match across all rows of seed~$j$ (decoder-cosine matching~\cite{bricken2023monosemanticity,lan2024universality}), a permutation-invariant statistic. The null is a \emph{subspace-conditioned} random-vector draw: real decoder directions concentrate in a subspace of effective rank $\mathrm{erank}{=}(\Sigma\sigma)^2/\Sigma\sigma^2{\approx}357$--$363$ (not the ambient $512$), so we draw null vectors inside each decoder's top-$\mathrm{erank}$ right-singular subspace and compare against the other decoder projected into it. This controls for the data-manifold concentration that an isotropic (full-$512$-d) null misses; the isotropic ratio is a reported lower bound.
\textbf{Stability (slot-wise, deprecated).} Cross-seed Jaccard of top-$k$ \emph{index} sets has analytical floor $k/(2D{-}k)$, but is \emph{floor-by-construction}: it is not invariant to the arbitrary per-seed feature permutation, so a single SAE whose features are relabelled by a random bijection scores ${\approx}0.0077$ on it (Sec.~\ref{sec:relabel}). We retain it only as an identical-permutation sanity check and report it explicitly deprecated.
\textbf{Naming.} Mean cosine of the assigned term, against a random-baseline of ${\approx}0.372$; we additionally report the fraction of features matched at $\geq0.7$ and $\geq0.9$, the concentration-robust signal (random directions cannot reach $0.9$).
\textbf{Faithfulness.} Point-biserial correlation of feature activations with in-distribution IU X-Ray MeSH/Problems labels, with a per-feature shuffle null. \textbf{LLM judge.} Each concept is scored Aligned/Unaligned/Uncertain against the reference report by a prompted LLM; faithfulness is reported as a \emph{LIFT} (method \%Aligned over a count-matched null's \%Aligned) to control for methods emitting different concept counts.

\section{Experiments and analysis}

\subsection{The baseline is a non-identifiable failure case (Gap 1)}
Table~\ref{tab:baseline} reports the parity baseline ($D{=}2048$, $k{=}32$, 8k steps, five seeds) on both corpora. Reconstruction is faithful (cosine $0.968$--$0.991$) yet cross-seed feature identity is absent: the permutation-invariant matched cosine is $0.30$--$0.33$, only $1.67$--$1.81{\times}$ the subspace-conditioned null, and the fraction of features matched at $\geq0.9$ is ${\leq}0.3\%$. Naming cosine is capped at $0.40$--$0.48$ with $0\%$ of features above $0.7$---the closest RadLex term to a live decoder row is barely closer than random. We read this as genuine non-identifiability of the learned factorisation, not a training bug (init distinct, decoders full-rank, no collapse). The deprecated slot-wise Jaccard lands at $0.0077$--$0.0084$, equal to its analytical floor---as expected, since it is floor-by-construction (Sec.~\ref{sec:relabel}).

\begin{table}[t]
\caption{Baseline Top-K SAE ($D{=}2048,k{=}32$) on IU X-Ray (${\sim}6$k) and ROCOv2 (${\sim}80$k, $10{\times}$). Matched-cosine is permutation-invariant; ``sub./iso.\ null'' ratios are observed best-match over the subspace-conditioned and isotropic nulls.}
\label{tab:baseline}
\small
\begin{tabular}{lcc}
\toprule
 & IU X-Ray (1$\times$) & ROCOv2 (10$\times$) \\
\midrule
Reconstruction cosine & 0.991 & 0.968 \\
Dead features (full test) & 15.0--17.8\% & 0.5--0.8\% \\
Matched cosine (obs) & 0.299 & 0.327 \\
Decoder erank & 363 & 357 \\
Subspace null ratio (honest) & 1.67$\times$ & 1.81$\times$ \\
Isotropic null ratio (lower bnd) & 1.98$\times$ & 2.16$\times$ \\
Frac.\ matched $\geq0.9$ & 0.0\% & 0.3\% \\
Naming cosine (live, max) & 0.40 & 0.48 \\
Slot-wise Jaccard (deprecated) & 0.0084 & 0.0077 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Relabeling control: the slot-wise metric is floor-by-construction}
\label{sec:relabel}
The slot-wise index Jaccard compares feature \emph{indices} across two SAEs with no alignment of which feature is which. Since independently seeded SAEs have no canonical feature ordering, the metric cannot distinguish ``two genuinely different SAEs'' from ``one SAE whose columns have been renamed.'' We verify this directly: we take one trained ROCOv2 SAE, permute its encoder/decoder rows by a random bijection~$\pi$ (a function-preserving relabelling---reconstruction cosine identical to $10^{-4}$), and measure both metrics between the original and the relabelled copy. Slot-wise Jaccard collapses to $0.0077$ (the chance floor, indistinguishable from real cross-seed $0.0077$), while the matched metric correctly reports identity ($1.0000$, frac${\geq}0.9{=}1.0$). This isolates permutation noise as the entire slot-wise signal: it is a binary identical-permutation check, not a stability/identifiability metric. All subsequent stability claims use the matched metric.

\subsection{Neither representation depth nor data scale restores identifiability}
Table~\ref{tab:patha} contrasts Path~A (768-d pre-projection hidden state) with the baseline. Operating on the 768-d hidden state markedly reduces dead features ($1.7\%$ vs.\ $16\%$) and improves naming cosine ($0.47$ vs.\ $0.42$, random $0.372$). Yet the permutation-invariant matched metric is unchanged in its verdict: both methods share a decoder subspace above the null (isotropic $1.97{\times}$/$2.63{\times}$, collapsing to ${\approx}1.3$--$1.5{\times}$ under the subspace-conditioned null), and almost no features match strongly ($0\%$ at $\geq0.9$). This is \emph{weak universality}~\cite{lan2024universality,leask2025canonical}: a shared subspace exists, but not canonical, reproducible features. Crucially, Path~A still operates on the \emph{pooled CLS token} (pre-projection, but still the contrastive-shaped global summary), so the representation-site mismatch is only partially addressed.

We then test the leading alternative explanation---data scale (Gap~4)---directly. Training the parity baseline on ROCOv2 (${\sim}80$k images, a $10{\times}$ increase over IU X-Ray) improves training health (dead features $16\%{\to}0.6\%$) but leaves identifiability flat (Table~\ref{tab:baseline}): matched cosine $0.299{\to}0.327$, subspace ratio $1.67{\times}{\to}1.81{\times}$, frac${\geq}0.9$ $0.0\%{\to}0.3\%$, naming capped at $0.40$--$0.48$. A $10{\times}$ corpus does not lift a single one of these above the non-identifiability regime. We conclude the binding constraint is \emph{not} data volume but the \emph{representation site}: SAE-on-CLS factorises a single, already-compressed, contrastive-shaped 512-d vector whose intrinsic dimension ($\mathrm{erank}{\approx}357$--$363$, participation rank measured at $3.8$--$56$) is far below the $2048$-way dictionary, so infinitely many equal-reconstruction tilings exist and no feature is canonical.

\begin{table}[t]
\caption{Baseline (512-d) vs.\ Path~A (768-d), parity configuration. Matched-cosine is permutation-invariant; ``obs/null'' is observed best-match cosine over the isotropic null.}
\label{tab:patha}
\small
\begin{tabular}{lcc}
\toprule
 & Baseline 512-d & Path A 768-d \\
\midrule
Slot-wise Jaccard & 0.0084 & 0.0083 \\
Analytical floor & 0.0079 & 0.0079 \\
Dead features (full test) & 16.0\% & 1.7\% \\
Naming cosine (live) & 0.42 & 0.47 \\
Matched cosine (obs) & 0.299 & 0.325 \\
Matched cosine (null) & 0.151 & 0.124 \\
obs/null & 1.97$\times$ & 2.63$\times$ \\
Frac.\ matched $\geq0.9$ & $\approx$0\% & $\approx$0\% \\
\bottomrule
\end{tabular}
\end{table}

A dictionary/$k$ sweep on Path~A (Table~\ref{tab:ablation}) confirms the pattern: a more conservative ($D{=}1024,k{=}16$) configuration shows the strongest subspace signal (matched obs/null $2.78{\times}$) and lowest naming noise, while aggressive overcompleteness ($D{=}4096,k{=}64$) dilutes it ($2.00{\times}$)---but no setting escapes weak universality.

\begin{table}[t]
\caption{Path~A dictionary/$k$ sweep (768-d, 8k steps).}
\label{tab:ablation}
\small
\begin{tabular}{lcccc}
\toprule
Preset & $D$ & $k$ & Dead & Matched obs/null \\
\midrule
Conservative & 1024 & 16 & 41.7\% & 2.78$\times$ \\
Default & 2048 & 32 & 12.9\% & 2.63$\times$ \\
Aggressive & 4096 & 64 & 6.6\% & 2.00$\times$ \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Path B is deterministic and judge-ready}
\label{sec:pathb}
SPLiCE decomposes every test image with a fixed, randomness-free solver---Orthogonal Matching Pursuit selects the $k{=}32$-atom support and non-negative least squares re-solves the coefficients on it---so the result is deterministic by construction and has no cross-seed instability. The output uses the SAE-compatible schema (\texttt{feature\_id}/\texttt{name}/\texttt{activation}), so the same judge reads SPLiCE concepts directly; this solver replaces an earlier post-hoc clamp that inflated reconstruction error $8$--$14{\times}$ (its run report still describes the superseded clamp). On IU X-Ray it covers all 1,515 test images in 9.5\,s (CPU) using 997/1,031 vocabulary terms (96.7\%), averaging 14.18 concepts/image (range 8--20); on ROCOv2 (${\sim}80$k) it covers all 15,958 images in 101\,s using 1,024/1,024 terms (100\%), averaging 8.56 concepts/image (range 3--15). Modality-gap correction raises the maximum atom--image cosine sharply on \emph{both} corpora (global maximum IU $0.49{\to}0.85$, ROCOv2 $0.54{\to}0.86$; per-image mean $0.42{\to}0.79$ and $0.40{\to}0.75$), confirming the gap rather than the solver was the binding issue. The frequent terms are mixed rather than clean clinical findings: ROCOv2 is dominated by imaging modalities (spiral CT, thoracic radiography, MR), reflecting its multi-anatomy corpus, while the IU vocabulary---a downloaded chest set---contains bilingual and off-topic entries (e.g.\ German translations such as \emph{Umbilikaler venöser Katheter}, and non-chest terms such as \emph{Backwash-Ileitis}); neither corpus yields the curated findings a hand-picked RadLex chest subset would, which is a property of the dictionaries supplied, not of SPLiCE itself.

\subsection{Concept organisation is dense but weakly separated}
Applied to all three IU sources (Table~\ref{tab:org}), the extension behaves as designed. SPLiCE, the densest source, groups 997 active concepts into 32 families, reducing concepts-per-image from 14.18 to 8.09 (redundancy reduction $1.75{\times}$), but with weak cosine silhouette ($0.020$); because the supplied IU vocabulary is bilingual and off-topic (Sec.~\ref{sec:pathb}), every cluster receives a RadLex ancestor yet the labels inherit the dictionary's noise (e.g.\ German \emph{Lunge}, \emph{Schmerz}, \emph{Kolitis} alongside \emph{vascular malformation}, \emph{compression fracture}). The rebuilt SAE baseline is no longer silent---it exposes 80 active concepts (9 families, silhouette $0.094$, zero empty images)---but these inherit the baseline's non-identifiability: the families are real groupings of noisy names, not stable concepts. Only the frozen Path~A hidden artifact remains degenerate (14 active concepts, 1{,}260/1{,}515 empty images). Replicating the extension on ROCOv2 SPLiCE gives 1{,}024 active concepts, 32 families, silhouette $0.066$, zero empty images---non-degenerate, confirming density scales with corpus size, though cluster separation stays weak across both corpora.

\begin{table}[t]
\caption{Concept organisation across sources.}
\label{tab:org}
\small
\begin{tabular}{lcccc}
\toprule
Source & Active & Clusters & Silhouette & Empty img.\ (/1515) \\
\midrule
SPLiCE & 997 & 32 & 0.020 & 0 \\
SAE baseline & 80 & 9 & 0.094 & 0 \\
SAE hidden & 14 & 4 & 0.284 & 1260 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{A small upper tail of features is genuinely faithful}
Despite global instability, a faithfulness ablation (historical $D{=}4096$ baseline) finds that $13.4\%$ of live features exceed the shuffle-null $95$th percentile in point-biserial correlation with IU X-Ray MeSH/Problems labels, and $1.2\%$ reach $|r|>0.30$ (strongest $|r|=0.405$, on a lung/hyperlucent feature). This is a real but small upper-tail signal atop an otherwise degenerate dictionary: concepts are simultaneously \emph{unstable cross-seed} (consensus shuffle-null $p=0.19$) and \emph{partially faithful} (Gap~2).

\subsection{The LLM-judge evaluation: machinery complete, inputs regenerated, run pending}
The judge is fully implemented and its mechanics are correct and reproducible: a verb$\to$label alias map (the prompt elicits SUPPORT/CONTRADICT/AMBIGUOUS verbs while the schema wants Aligned/Unaligned/Uncertain, so verbs are normalised), $\mathrm{temperature}{=}0$ with a global seed for determinism, infrastructure-error pairs excluded from the resume dedup so transient failures are retried, and the faithfulness metric is a count-matched \emph{LIFT} (method \%Aligned over a null emitting the same concept count) persisted per method. The per-image pseudo-reports feeding the judge were regenerated cleanly for all nine method/dataset combinations (IU X-Ray: baseline, SPLiCE, null, null$_{k5}$, sae\_hidden; ROCOv2: baseline, SPLiCE, null, null$_{k5}$)---$1{,}515$ and $15{,}958$ images respectively, real image IDs, schema-compatible across methods. The single remaining step is a converged GPU execution (MedGemma-4B~\cite{sellergren2025medgemma} on a T4) over these inputs, scheduled as separate work. We therefore report the method-level LIFT ranking as \emph{not yet measured}; partial baseline-only runs across models disagree widely (MedGemma-4B $84.5\%$ Aligned, Llama-3.1-8B~\cite{meta2024llama3} $0\%$, a 26B Gemma~\cite{gemma2024} $4\%$), underscoring that absolute Aligned-rates are judge-dependent and only the within-judge LIFT against the count-matched null is interpretable.

\section{Conclusions}
We set out to discover interpretable concepts in a medical VLM without supervision, and instead produced a rigorous characterisation of where and why this fails. Two methodological findings reframe the problem before any method comparison: the slot-wise cross-seed Jaccard used as the stability headline is \emph{floor-by-construction} (a relabelled copy of one SAE scores ${\approx}0.0077$ on it, proved by control), so non-identifiability must be measured permutation-invariantly; and under the resulting matched metric with a subspace-conditioned null, SAE concepts are genuinely non-identifiable (best-match $1.67$--$1.81{\times}$ the null, ${\approx}0\%$ at $\geq0.9$, naming capped at $0.40$--$0.63$). We then localised the cause: neither moving the SAE to the 768-d pre-projection hidden state (Path~A) nor a $10{\times}$ increase in data (ROCOv2) lifts identifiability---the binding constraint is the \emph{representation site}, not data volume. SAE-on-CLS factorises a single, already-compressed, contrastive-shaped global vector over a $30$--$500{\times}$ overcomplete dictionary, so no canonical features exist. SPLiCE, operating on a fixed dictionary, sidesteps the estimation problem and is the sole deterministic, judge-ready alternative. A concept-organisation extension still reduces redundancy in SPLiCE output ($1.75{\times}$, 32 families), though cluster separation is weak (silhouette $0.020$) and the family labels inherit the noise of the supplied dictionary, and a faithfulness ablation shows a small (${\sim}13\%$) but real upper tail of clinically-correlated features atop an otherwise degenerate dictionary.

Three lessons generalise beyond our setting. First, interpretability claims about discovered concepts must be calibrated against analytical nulls \emph{and} a permutation-invariant statistic---slot-wise index overlap is degenerate and its floor can be mistaken for a result. Second, when a learned factorisation fails, data scale is the suspect of last resort: a $10{\times}$ corpus changed training health but not identifiability, because the failure was in the input representation. Third, an honest evaluation---including a metric artefact and a pending judge run---is more scientifically useful than a curated one.

\textbf{Limitations and future work.} The judge LIFT ranking must be completed (machinery and regenerated inputs are ready; a converged GPU run is the remaining step) with bootstrap confidence intervals. The decisive untested mitigation is the representation site itself: an SAE on the BiomedCLIP \emph{patch-token residual stream} at a mid layer (not the pooled CLS), following the SAE-interpretability literature~\cite{bricken2023monosemanticity}; this is the experiment our diagnosis predicts would restore identifiability.

%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{biblio}


\end{document}
\endinput
%%
